# (PART) DEMO DATASET {-}

# Microbiome Demo Datasets {#demo-datasets}

In this section, we provide demo datasets for microbiome analysis. These datasets have been preprocessed and transformed in previous sessions, and now we are preparing to analyze them. We begin by importing the R and Phyloseq objects.

```{r, include=FALSE}
knitr::opts_chunk$set(
  echo  =TRUE,
  message  =FALSE,
  warning  =FALSE,
  cache  =FALSE,
  fig.path = "figures",
  comment  =NA)
  
```


```{bash eval=FALSE, include=FALSE}
mkdir -p figures
mkdir -p data

cp -r ../imap-data-preparation/data/* data/

```

## Import preprepared data
```{r message=FALSE, warning=FALSE}
# Load required packages
library(phyloseq)
library(tidyverse)

cat("\nSaved RData objects\n\n")

load("data/ps_transformed.rda", verbose = T)
load("data/bray_distances.rda", verbose = T)
load("data/reduced_dimension.rda", verbose = T)
load("data/phyloseq_raw_rel_psextra_df_objects.rda", verbose = T)

```


## Demo data for QIIME2 analysis

The demo data for QIIME2 analysis is downloaded and stored in the `data/qiime2` directory. You can use these files to practice QIIME2 workflows and analysis techniques.

The data was sourced from the QIIME2 tutorials available at [https://docs.qiime2.org/2024.4/data/tutorials/moving-pictures/](https://docs.qiime2.org/2018.4/data/tutorials/moving-pictures/).

It contains the following files:

- Sample metadata
- Feature table
- Taxonomy
- Shannon vector
- Rooted tree

This data is essential for leveraging downstream analysis aimed at understanding data analyzed using the QIIME2 pipeline.


```R
if (!dir.exists("data")){dir.create("data")}
if (!dir.exists("data/qiime2")){dir.create("data/qiime2")}

download.file("https://docs.qiime2.org/2024.2/data/tutorials/moving-pictures/table.qza", "data/qiime2/feature_table.qza")
download.file("https://data.qiime2.org/2024.2/tutorials/moving-pictures/sample_metadata.tsv", "data/qiime2/sample_metadata.tsv")
download.file("https://docs.qiime2.org/2024.2/data/tutorials/moving-pictures/taxonomy.qza", "data/qiime2/taxonomy.qza")
download.file("https://docs.qiime2.org/2024.2/data/tutorials/moving-pictures/rooted-tree.qza", "data/qiime2/rooted_tree.qza")
download.file("https://docs.qiime2.org/2024.2/data/tutorials/moving-pictures/core-metrics-results/shannon_vector.qza", "data/qiime2/shannon_vector.qza")

```

