# (PART) ORDINATION EXPLORATION{-}

# Ordination of Abundance Data

```{r include=FALSE}
library(vegan)
library(ape)
library(dplyr)
library(microViz)

```

## Import count data
```{r}
load("../imap-data-preparation/data/external/phyloseq_objects.rda", verbose = TRUE)
ps <- dietswap %>% tax_fix() %>% phyloseq_validate()
ps

cat("\n\n")

otu_table(ps)[1:10,1:8]
```

## Interactive ordination
- Package: microViz
- Function: ord_explore(phyloseq object)
- Reference: https://github.com/david-barnett/microViz


## Static ordination
For simplicity, we will demonstrate static ordinations generated using different techniques.

```{r}
library(microViz)

ps <- ps_dietswap %>% 
  tax_fix() %>% 
  phyloseq_validate()


```


## Constrained partial ordination
You could visualise the effect of the (numeric/logical) variables in your permanova directly using the ord_plot function with constraints (and conditions).

```{r}
perm2 <- aitchison_dists %>%
  dist_permanova(variables = c("weight", "african", "sex"), seed = 321)
```


```{r}
ps %>%
 tax_transform(rank = "Genus", trans = "clr") %>%
 dist_calc(dist = "bray") %>%
 ord_calc(
  method = "PCoA"
 ) %>% 
 ord_plot(
  axes = c(1, 2),
  plot_taxa = 1:3,
  colour = "azure4", fill = "azure4",
  shape = "circle", alpha = 0.5,
  size = 2
 )
```


## Static Ordination
```{r}

```


### PCA: Principal component analysis
```{r PCA_ord}
## Phylum-level

p_pca_ord <- ps %>%
 tax_transform(rank = "Phylum", trans = "identity") %>%
 ord_calc(method = "PCA") %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Platform", fill = "Platform",
  shape = "Platform", alpha = 0.9,
  size = 3) + 
 scale_shape_girafe_filled() #+
  labs(caption = NULL)
p_pca_ord +
    theme_bw()

## Class-level
c_pca_ord <- ps %>%
 tax_transform(rank = "Class", trans = "identity") %>%
 ord_calc(method = "PCA") %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Platform", fill = "Platform",
  shape = "Platform", alpha = 0.9,
  size = 3) + 
 scale_shape_girafe_filled() #+
  labs(caption = NULL)
c_pca_ord +
    theme_bw()

## Order-level
o_pca_ord <- ps %>%
 tax_transform(rank = "Order", trans = "identity") %>%
 ord_calc(method = "PCA") %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Platform", fill = "Platform",
  shape = "Platform", alpha = 0.9,
  size = 3) + 
 scale_shape_girafe_filled() #+
  labs(caption = NULL)
o_pca_ord +
    theme_bw()

## Family-level
f_pca_ord <- ps %>%
 tax_transform(rank = "Family", trans = "identity") %>%
 ord_calc(method = "PCA") %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Platform", fill = "Platform",
  shape = "Platform", alpha = 0.9,
  size = 3) + 
 scale_shape_girafe_filled() #+
  labs(caption = NULL)
f_pca_ord +
    theme_bw()
  
## Genus-level
g_pca_ord <- ps %>%
 tax_transform(rank = "Genus", trans = "identity") %>%
 ord_calc(method = "PCA") %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Platform", fill = "Platform",
  shape = "Platform", alpha = 0.9,
  size = 3) + 
 scale_shape_girafe_filled() #+
  labs(caption = NULL, title = "Genus-level NMDS")
g_pca_ord +
    theme_bw()
  
## Species-level
s_pca_ord <- ps %>%
 tax_transform(rank = "Species", trans = "identity") %>%
 ord_calc(method = "PCA") %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Platform", fill = "Platform",
  shape = "Platform", alpha = 0.9,
  size = 3) + 
 scale_shape_girafe_filled() #+
  labs(caption = NULL, title = "Species-level PCA")
s_pca_ord +
    theme_bw()

```

### PCoA: Principal Coordinate Analysis
```{r PCoA_ord}
## Phylum-level
p_pcoa_ord <- ps %>%
 tax_transform(rank = "Phylum", trans = "identity") %>%
 dist_calc(dist = "bray") %>%
 ord_calc(method = "PCoA") %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Platform", fill = "Platform",
  shape = "Platform", alpha = 0.9,
  size = 3) + 
 scale_shape_girafe_filled() #+
  labs(caption = NULL)
p_pcoa_ord +
    theme_bw()

## Class-level
c_pcoa_ord <- ps %>%
 tax_transform(rank = "Class", trans = "identity") %>%
 dist_calc(dist = "bray") %>%
 ord_calc(method = "PCoA") %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Platform", fill = "Platform",
  shape = "Platform", alpha = 0.9,
  size = 3) + 
 scale_shape_girafe_filled() #+
  labs(caption = NULL)
c_pcoa_ord +
    theme_bw()

## Order-level
o_pcoa_ord <- ps %>%
 tax_transform(rank = "Order", trans = "identity") %>%
 dist_calc(dist = "bray") %>%
 ord_calc(method = "PCoA") %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Platform", fill = "Platform",
  shape = "Platform", alpha = 0.9,
  size = 3) + 
 scale_shape_girafe_filled() #+
  labs(caption = NULL)
o_pcoa_ord +
    theme_bw()

## Family-level
f_pcoa_ord <- ps %>%
 tax_transform(rank = "Family", trans = "identity") %>%
 dist_calc(dist = "bray") %>%
 ord_calc(method = "PCoA") %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Platform", fill = "Platform",
  shape = "Platform", alpha = 0.9,
  size = 3) + 
 scale_shape_girafe_filled() #+
  labs(caption = NULL)
f_pcoa_ord +
    theme_bw()
  
## Genus-level
g_pcoa_ord <- ps %>%
 tax_transform(rank = "Genus", trans = "identity") %>%
 dist_calc(dist = "bray") %>%
 ord_calc(method = "PCoA") %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Platform", fill = "Platform",
  shape = "Platform", alpha = 0.9,
  size = 3) + 
 scale_shape_girafe_filled() #+
  labs(caption = NULL, title = "Genus-levell PCoA")
g_pcoa_ord +
    theme_bw()
  
## Species-level
s_pcoa_ord <- ps %>%
 tax_transform(rank = "Species", trans = "identity") %>%
 dist_calc(dist = "bray") %>%
 ord_calc(method = "PCoA") %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Platform", fill = "Platform",
  shape = "Platform", alpha = 0.9,
  size = 3) + 
 scale_shape_girafe_filled() #+
  labs(caption = NULL, title = "Species-level PCoA")
s_pcoa_ord +
    theme_bw()

```

### NMDS: Non-Metric Multidimensional Scaling
```{r NMDS_ord}
## Phylum-level
p_nmds_ord <- ps %>%
 tax_transform(rank = "Phylum", trans = "identity") %>%
 dist_calc(dist = "bray") %>%
 ord_calc(method = "NMDS") %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Platform", fill = "Platform",
  shape = "Platform", alpha = 0.9,
  size = 3) + 
 scale_shape_girafe_filled() #+
  labs(caption = NULL)
p_nmds_ord +
    theme_bw()

## Class-level
c_nmds_ord <- ps %>%
 tax_transform(rank = "Class", trans = "identity") %>%
 dist_calc(dist = "bray") %>%
 ord_calc(method = "NMDS") %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Platform", fill = "Platform",
  shape = "Platform", alpha = 0.9,
  size = 3) + 
 scale_shape_girafe_filled() #+
  labs(caption = NULL)
c_nmds_ord +
    theme_bw()

## Order-level
o_nmds_ord <- ps %>%
 tax_transform(rank = "Order", trans = "identity") %>%
 dist_calc(dist = "bray") %>%
 ord_calc(method = "NMDS") %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Platform", fill = "Platform",
  shape = "Platform", alpha = 0.9,
  size = 3) + 
 scale_shape_girafe_filled() #+
  labs(caption = NULL)
o_nmds_ord +
    theme_bw()

## Family-level
f_nmds_ord <- ps %>%
 tax_transform(rank = "Family", trans = "identity") %>%
 dist_calc(dist = "bray") %>%
 ord_calc(method = "NMDS") %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Platform", fill = "Platform",
  shape = "Platform", alpha = 0.9,
  size = 3) + 
 scale_shape_girafe_filled() #+
  labs(caption = NULL)
f_nmds_ord +
    theme_bw()
  
## Genus-level
g_nmds_ord <- ps %>%
 tax_transform(rank = "Genus", trans = "identity") %>%
 dist_calc(dist = "bray") %>%
 ord_calc(method = "NMDS") %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Platform", fill = "Platform",
  shape = "Platform", alpha = 0.9,
  size = 3) + 
 scale_shape_girafe_filled() #+
  labs(caption = NULL, title = "Genus-level NMDS")
g_nmds_ord +
    theme_bw()
  
## Species-level
s_nmds_ord <- ps %>%
 tax_transform(rank = "Species", trans = "identity") %>%
 dist_calc(dist = "bray") %>%
 ord_calc(method = "NMDS") %>% 
 ord_plot(
  axes = c(1, 2),
  colour = "Platform", fill = "Platform",
  shape = "Platform", alpha = 0.9,
  size = 3) + 
 scale_shape_girafe_filled() #+
  labs(caption = NULL, title = "Species-level NMDS")
s_nmds_ord  +
    theme_bw()
    
```

```{r ordination_gif, echo=FALSE, animation.hook='gifski', interval=3, fig.path="./GIFS/"}
g_pca_ord
s_pca_ord
g_pcoa_ord
s_pcoa_ord
g_nmds_ord
s_nmds_ord

```

# Ordination Methods Explained

```{block, type="infoicon", echo=TRUE}
**Summary of ordination methods**

1. PCA (Principal Component Analysis): PCA can be used to identify the major patterns of variation in taxa abundance data, revealing the primary axes of variability.
2. PCoA (Principal Coordinates Analysis): PCoA is employed to visualize the similarity or dissimilarity in taxa abundance patterns, representing them in a lower-dimensional space.
3. MDS (Multidimensional Scaling): MDS helps visualize the dissimilarity or similarity in taxa abundance data, providing insights into the relationships between samples or communities.
4. NMDS (Non-Metric Multidimensional Scaling): NMDS is a variation of MDS that focuses on preserving the rank order of dissimilarities in taxa abundance, aiding in the visualization of complex community data.
5. DPCoA (Double Principal Coordinate Analysis): DPCoA allows for the visualization of relationships between multiple dissimilarity matrices in taxa abundance data.
6. DCA (Detrended Correspondence Analysis): DCA captures the major gradients of variation in taxa abundance data, facilitating the exploration and interpretation of community composition.
7. CCA (Canonical Correspondence Analysis): CCA examines the associations between taxa abundance and environmental variables, identifying the environmental gradients influencing the community structure.
8. RDA (Redundancy Analysis): RDA extends CCA by quantifying the proportion of variation in taxa abundance explained by environmental variables and shared effects.
9. CAP (Canonical Analysis of Principal coordinates): CAP combines ordination and classification techniques to assess the separation of predefined groups based on taxa abundance data.
10. CA (Correspondence Analysis): CA visualizes the associations between taxa and different categories, aiding in the exploration of patterns in ecological community data.
```

## Load data for Beta diversity analysis
```{r ordi_methods}
load("data/ps_transformed.rda", verbose = TRUE)
load("data/bray_distances.rda", verbose = TRUE)
```

## PCA: Principal Component Analysis
- RDA (redundancy analysis) method in vegan package performs principal components analysis (PCA).

```{r pca-ordi}
library(tidyverse)
library(phyloseq)
set.seed(23)


metadata <- data.frame(sample_data(ps_raw)) %>%
  tibble::rownames_to_column("sample_id") %>% 
  rename(bmi=bmi_group)

ps_pca = phyloseq::ordinate(ps_raw, "RDA", "euclidean")
# Create scree plot with plot_scree()
phyloseq::plot_scree(ps_pca) + 
  geom_col(fill = "steelblue") + 
  theme_bw() +
  labs(x = "Number of Components", y = "Proportion of Variance Explained", subtitle = "PCA scree plot for all axes") +
  theme(axis.text.x = element_blank()) +
  scale_y_continuous(labels = percent)

eigenvalues <- ps_pca$CA$eig[1:10]

# Create a dataframe with eigenvalues and axis numbers
pca_axes <- data.frame(PC = 1:length(eigenvalues), Eigenvalue = eigenvalues)

# Create the bar plot and scree plot with ggplot2 - axis 1 to 10
pca_axes %>% 
  mutate(Eigenvalue = Eigenvalue/sum(Eigenvalue)) %>% 
ggplot(aes(x = round(PC,4), y = Eigenvalue)) +
  geom_col(fill = "steelblue") +
  geom_line(color = "black", size = 1) +   
  geom_point(color = "#222222", size = 3) +
  labs(x = "Number of Components", y = "Eigen Value", subtitle = "PCA scree plot of top 10 significant PC") +
  theme_test() +
  scale_y_continuous(labels = percent)


# - Extract PCoA vectors.
# - Join metadata and the PCoA data.
# - Plot the first two axes.

pca_samples <- as.data.frame(ps_pca$CA$u) %>% 
  tibble::rownames_to_column("sample_id")

metadata_pcoa <- inner_join(metadata, pca_samples, by='sample_id')

metadata_pcoa %>% 
  ggplot(aes(x=round(PC1, 4), y=round(PC2,4), color=nationality)) +
	geom_point(shape=19, size=2) +
	# coord_fixed() +
	labs(subtitle="RDA-PCA ordination using Bray-Curtis distance dissimilarities \nOrdination method: phyloseq::geom_point()",
		x="PC 1",
		y="PC 2",
		color = "Nationality") +
	theme_test() +
  scale_y_continuous(labels = percent)

```



## PCoA: Principal Coordinate Analysis {#pcoa_analysis}
```{block, type="infoicon", echo=TRUE}
- PCoA is Classical Metric Multidimensional Scaling (MDS).
- All variables are scaled to unit variance before the analysis.
- PCoA enables us to plot two axes against each other.
- The PCoA scree plot plots the proportion of variance explained by each principal coordinate on the y-axis against the corresponding coordinate number on the x-axis. 
    - Axis 1 explains the most variation in the data
    - Axis 2 explains the second most variation, and so forth.
    - It helps visualize the diminishing returns of adding more coordinates
    - Identifies the significant axes that capture most of the variation.
```


```{r pcoa_ordi, message=FALSE, warning=FALSE}
library(tidyverse)

metadata <- data.frame(sample_data(ps_raw)) %>%
  tibble::rownames_to_column("sample_id") %>% 
  rename(bmi=bmi_group)

# Perform PCoA on the bray dissimilarity data
ps_pcoa = phyloseq::ordinate(ps_log10p, "PCoA", "bray", weighted=TRUE)

# Create scree plot with plot_scree()
phyloseq::plot_scree(ps_pcoa) + 
  geom_col(fill = "steelblue") + 
  theme_bw() +
  labs(x = "Number of Coordinates", y = "Percent of Variance Explained", subtitle = "PCoA scree plot of all axes") +
  theme(axis.text.x = element_blank()) +
  scale_y_continuous(labels = percent)

eigenvalues <- ps_pcoa$values$Eigenvalues[1:10]

# Create a dataframe with eigenvalues and axis numbers
df <- data.frame(Axis = 1:length(eigenvalues), Eigenvalue = eigenvalues)

# Create the bar plot and scree plot with ggplot2 - axis 1 to 10
ggplot(df, aes(x = Axis, y = Eigenvalue)) +
  geom_col(fill = "steelblue") +
  geom_line(color = "black", size = 1) +   
  geom_point(color = "#222222", size = 3) +
  labs(x = "Number of Coordinates", y = "Percent of Variance Explained", subtitle = "PCoA scree plot of top 10 significant axes") +
  theme_test() +
  scale_y_continuous(labels = percent)


# - Extract PCoA vectors.
# - Join metadata and the PCoA data.
# - Plot the first two axes.

pcoa_vectors <- as.data.frame(ps_pcoa$vectors) %>% 
  rename_all(~ str_replace(., "\\.", "")) %>% 
  tibble::rownames_to_column("sample_id")

metadata_pcoa <- inner_join(metadata, pcoa_vectors, by='sample_id')

metadata_pcoa %>% 
  ggplot(aes(x=Axis1, y=Axis2, color=nationality)) +
	geom_point(shape=19, size=2) +
	# coord_fixed() +
	labs(subtitle="Ex 1: PCoA ordination using Bray-Curtis distance dissimilarities \nOrdination method: phyloseq::ordinate()",
		x="PCoA Axis 1",
		y="PCoA Axis 2",
		color = "Nationality") +
	theme_test()



## Using PCoA data from cmdscale() function

set.seed(2022)

dist <- ps_log10p_bray_dist

pcoa_cmd <- cmdscale(dist, k = (nrow(metadata) - 1), eig = TRUE)
pcoa_variance <- data.frame(PCO = paste("PCO",  seq(1:nrow(as.matrix(ps_log10p_bray_dist))), sep = ""), 
                           ExplainedVar = round(pcoa_cmd$eig^2/sum(pcoa_cmd$eig^2), 4))
### PCoA with formatted y-axis
pcoa_1_2 <- data.frame(PCO1 = pcoa_cmd$points[,1], PCO2 = pcoa_cmd$points[,2]) %>% 
  tibble::rownames_to_column("sample_id") %>% 
  inner_join(metadata, pcoa_1_2, by = "sample_id")

perc_expned <- format(round((100 * pcoa_cmd$eig / sum(pcoa_cmd$eig)), digits =1), nsmall=1, trim=TRUE)

library(glue)
labels <- c(glue("PCoA Axis 1 ({perc_expned[1]}%)"),
            glue("PCoA Axis 2 ({perc_expned[2]}%)"))

pcoa_1_2 %>% 
ggplot(aes(x = PCO1, y = PCO2, color = nationality)) +
	geom_point(shape=19, size=2) +
	# coord_fixed() +
	labs(subtitle="Ex 2: PCoA ordination using Bray-Curtis distance dissimilarities \nOrdination method: cmdscale()",
		x="PCoA Axis 1",
		y="PCoA Axis 2",
		color = "Nationality") +
	theme_test()


# With more aesthetic parameters.
pcoa_vectors %>%
  select(sample_id, Axis1, Axis2) %>%
  left_join(metadata) %>%
  left_join(readRDS("data/df_alpha_wide.rds")) %>%
  ggplot(aes(x=Axis1, y=Axis2, color=nationality, shape=sex, size=shannon)) +
  geom_point(alpha=0.5) + 
  theme_q2r() +
  scale_shape_manual(values=c(16,1), name="Sex") + 
  scale_size_continuous(name="Shannon Diversity") +
  scale_color_discrete(name="Nationality") +	
  labs(subtitle="PCoA ordination with additional aesthetic parameters.",
		x="PCoA1",
		y="PCoA2",
		color = "Nationality") +
	theme_test()



library(microbial)
plotbeta(ps_raw, group="nationality")
```


## NMDS: Nonmetric Multidimensional Scaling {#nmds-ordi}
```{block, type="infoicon", echo=TRUE}
NMDS plots show the samples as points in the reduced space.

- The proximity of points on the NMDS plot reflects their similarity or dissimilarity, with closer points representing more similar samples or objects. 
- The NMDS plot can be further annotated with metadata to aid in the interpretation of the patterns.
```


```{r nmds_ordi}
### Ordination
set.seed(25)

ps <- ps_log10p
ps_nmds = phyloseq::ordinate(ps, "NMDS", "bray", weighted=TRUE)
nmds_points <- as.data.frame(ps_nmds$points) %>% 
  tibble::rownames_to_column("sample_id")

metadata_nmds <- inner_join(metadata, nmds_points, by='sample_id')

metadata_nmds %>% 
  ggplot(aes(x=MDS1, y=MDS2, color=nationality)) +
  geom_point(size=2, alpha=0.75, shape=19, ) + 
  scale_colour_brewer(type="qual", palette="Set1") + 
	# coord_fixed() +
	labs(subtitle="NMDS ordination using Bray-Curtis distance dissimilarities \nOrdination method: phyloseq::ordinate()",
		x="NMDS 1",
		y="NMDS 2",
		color = "Nationality") +
    geom_point(size=2) +
	theme_test()


nmds_points %>%
  select(sample_id, NMDS1=MDS1, NMDS2=MDS2) %>%
  left_join(metadata) %>%
  left_join(readRDS("data/df_alpha_wide.rds")) %>%
  ggplot(aes(x=NMDS1, y=NMDS2, color=nationality, shape=sex, size=shannon)) +
  geom_point(alpha=0.5) + 
  theme_q2r() +
  scale_shape_manual(values=c(16,1), name="Sex") + 
  scale_size_continuous(name="Shannon Diversity") +
  scale_color_discrete(name="Nationality") +	
  labs(subtitle="NMDS ordination with additional aesthetic parameters.",
		x="NMDS1",
		y="NMDS2",
		color = "Nationality") +
	theme_test()


### NMDS Stress Plot

stressplot(ps_nmds)

```



## DPCoA: Double Principal Coordinate Analysis {#dpcoa}
```{block, type="infoicon", echo=TRUE}
DPCoA requires `asymmetric dissimilarity matrices`, where the dissimilarity value between Sample A and Sample B can be different from the dissimilarity value between Sample B and Sample A. 

- DPCoA is not applicable to `symmetric dissimilarity` measures like Bray-Curtis.
- Weighted UniFrac dissimilarities are asymmetric
- DPCoA can be applied to visualize and analyze the relationships between samples based on these dissimilarities. 
- DPCoA can help reveal patterns, clusters, or gradients in the weighted UniFrac dissimilarity structure
```

```{r}
phy_tree <- rtree(ntaxa(ps_dietswap))
dietswap <- merge_phyloseq(dietswap, phy_tree)
```


```{r DPCoA_ordi}
ps <- ps_log10p
ps_dpcoa = phyloseq::ordinate(ps, "DPCoA", "wunifrac")

# Create scree plot with plot_scree()
phyloseq::plot_scree(ps_dpcoa) + 
  geom_col(fill = "steelblue") + 
  theme_bw() +
  labs(x = "Number of Coordinates", y = "Percent of Variance Explained", subtitle = "PCoA scree plot of all axes") +
  theme(axis.text.x = element_blank()) +
  scale_y_continuous(labels = percent)

eigenvalues <- ps_dpcoa$eig[1:10]

# Create a dataframe with eigenvalues and axis numbers
df <- data.frame(Axis = 1:length(eigenvalues), Eigenvalue = eigenvalues)

# Create the bar plot and scree plot with ggplot2 - axis 1 to 10
ggplot(df, aes(x = Axis, y = Eigenvalue)) +
  geom_col(fill = "steelblue") +
  geom_line(color = "black", size = 1) +   
  geom_point(color = "#222222", size = 3) +
  labs(x = "Number of Double Coordinates", y = "Percent of Variance Explained", subtitle = "PCoA scree plot of top 10 significant axes") +
  theme_test() +
  scale_y_continuous(labels = percent)


# Ordination 

dpcoa_tab <- as.data.frame(ps_dpcoa$tab) %>% 
  tibble::rownames_to_column("sample")
ps <- ps_log10p

metadata <- data.frame(sample_data(ps)) %>% 
  select(-sample) %>% 
  tibble::rownames_to_column("sample")

metadata_nmds <- inner_join(metadata, dpcoa_tab, by='sample')

metadata_nmds %>% 
  ggplot(aes(x=A1, y=A2, color=nationality)) +
	geom_point(shape=19, size=2) +
	# coord_fixed() +
	labs(subtitle="DPCoA ordination using Weighted UniFrac distances\nOrdination method: cmdscale()",
		x="DPCoA Axis 1",
		y="DPCoA Axis 2",
		color = "Nationality") +
	theme_test()

```




## Grouped ordination plots
```{r grouped_ordi, fig.height=10, fig.width=10}

library(phyloseq)
library(plyr)
library(ggpubr)

dist = "bray"
ord_meths = c( "RDA", "PCoA", "MDS", "NMDS", "DCA", "CCA")

plist <- llply(ord_meths, function(i, physeq, dist) {
  ordi_plot <- phyloseq::ordinate(physeq, method = i, distance = dist)
  plot_ordination(
    physeq, ordi_plot, "samples",
    color = "nationality") +
    theme(text = element_text(size = 8)) +
    labs(color = "Nationality", subtitle = i) +
    geom_point(size = 2) +
    theme_test()+
    theme(plot.subtitle = element_text(face = "bold"))

}, physeq = ps_log10p, dist = dist)
 

library(ggpubr)

ggpubr::ggarrange(plist[[1]], 
          plist[[2]], 
          plist[[3]], 
          plist[[4]], 
          plist[[5]], 
          plist[[6]], 
          ncol = 2, nrow = 3, common.legend = TRUE, legend = "right", labels = NULL )



# Add polygons


dist = "bray"
ord_meths = c( "CCA", "DCA", "MDS", "NMDS", "PCoA", "RDA")
names(plist) <- ord_meths
pdataframe = ldply(plist, function(x){
    df = x$data[, 1:2]
    colnames(df) = c("Axis_1", "Axis_2")
    return(cbind(df, x$data))
})
names(pdataframe)[1] = "method"
pdataframe %>% 
  ggplot(aes(Axis_1, Axis_2, color=nationality, fill=nationality)) +
  geom_point(size=2) + 
  geom_polygon() +
  facet_wrap(method ~., scales="free", ncol = 2) +
  scale_fill_brewer(type="qual", palette="Set2") +
  scale_colour_brewer(type="qual", palette="Set2") + 
  theme_test()
```